---
title: '2021 Census: population and household change in Leicester'
author: "Stefano De Sabbata"
date: '`r lubridate::now()`'
output: 
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libraries, warning=FALSE, message=FALSE}
library(tidyverse)
library(magrittr)
library(nomisr)
library(skimr)
library(ggrepel)
library(ggExtra)
require(scales)
library(classInt)
library(sf)
library(ggmap)
library(ggsflabel)
library(mapsf)
```


This document uses the [Demography and migration data](https://www.ons.gov.uk/peoplepopulationandcommunity/populationandmigration/populationestimates/articles/demographyandmigrationdatacontent/2022-11-02#demography-unrounded-population-estimates) for England and Wales from the [Census 2021](https://census.gov.uk/) released on 2 November 2022, as well as [Output Areas (December 2021) Boundaries Generalised Clipped EW (BGC)](https://geoportal.statistics.gov.uk/datasets/ons::output-areas-december-2021-boundaries-generalised-clipped-ew-bgc/about) and the [OAs to LSOAs to MSOAs (2021) to Local Enterprise Partnership to LAD (May 2022) Lookup in England](https://geoportal.statistics.gov.uk/datasets/ons::oas-to-lsoas-to-msoas-2021-to-local-enterprise-partnership-to-lad-may-2022-lookup-in-england-/about). Source: [Office for National Statistics](https://www.ons.gov.uk/) licensed under the [Open Government Licence v3.0](https://www.nationalarchives.gov.uk/doc/open-government-licence/version/3/). Contains both Ordnance Survey and ONS Intellectual Property Rights. Contains OS data © Crown copyright and database right 2022.


## Data

The 2021 Census data have been downloaded to the `storage` folder using the scripts `002-download-data-2022-11-02.R` and `003-download-oa-geometries.R` available in the `000-data` folder.

### Load data

Load 2021 Census data from the `storage` folder.

```{r load_data}
population_density_2021 <- read_csv(
    "../storage/atc-ts-demmig-ur-pd-oa-oa.csv"
  ) %>% 
  rename(
    OA21CD = `Output Areas Code`,
    OA21NM = `Output Areas Label`,
    population_density = `Population Density`
  )

oa_2021_geo <-
  st_read("../storage/Leicester_2021_OAs.geojson")

population_2021_est <-
  oa_2021_geo %>% 
  st_transform(crs = 27700) %>% 
  mutate(
    area = st_area(geometry) %>% as.numeric()
  ) %>% 
  st_drop_geometry() %>% 
  left_join(
    population_density_2021
  ) %>% 
  mutate(
    population = as.numeric(round(population_density * (area / 1000000)))
  ) %>% 
  select(OA21CD, population_density, area, population)

households_2021 <- read_csv(
    "../storage/atc-ts-demmig-hh-ct-oa-oa.csv"
  ) %>% 
  rename(
    OA21CD = `Output Areas Code`,
    OA21NM = `Output Areas Label`,
    households = `Count`
  )
```

```{r pop_household_join}
pop_hh_2021 <-
  oa_2021_geo %>% 
  left_join(
    population_2021_est
  ) %>% 
  left_join(
    households_2021
  ) %>% 
  mutate(
    pop_hh_2021 = population / households
  )
```

```{r get_2011_census_data}
pop_hh_2011 <-
  st_read("../storage/Leicester_2011_OAs_population_households.geojson") %>% 
  mutate(
    pop_hh_2011 = all_persons_2011 / households_2011
  )
```

### Check density estimation

Checking for outliers in the density estimation.

```{r}
popdens_area_model <-
  population_2021_est %>% 
  lm(
    log10(population_density) ~
      log10(area),
    data = .
  )

popdens_area_model %>% 
  summary()

population_2021_est_lm <-
  population_2021_est %>% 
  mutate(
    cooksd =
      popdens_area_model %>% 
      cooks.distance()
  )
```


```{r}
population_2021_est_lm %>% 
  mutate(
    plot_label = if_else(
      cooksd > 0.01, 
      OA21CD, NA_character_
    )
  ) %>% 
  ggplot(
    aes(
      x = area,
      y = population_density,
      colour = cooksd,
      label = plot_label
    )
  ) +
  scale_color_viridis_c(option = "inferno", direction = -1) +
  geom_point() +
  geom_text_repel() +
  scale_x_log10() +
  scale_y_log10() +
  theme_bw()
```

```{r, echo=FALSE, message=FALSE, warning=FALSE, results='hide', fig.height = 8, fig.width = 8}
pop_hh_2021 %>% 
  left_join(
    population_2021_est_lm %>% 
      mutate(
        plot_label = if_else(
          cooksd > 0.01, 
          OA21CD, NA_character_
        )
      )
  ) %>% 
  ggplot() +
  geom_sf(
    mapping = aes(
      fill = plot_label,
      geometry = geometry
    ),
    inherit.aes = FALSE,
    color = NA,
    alpha = 0.8
  ) +
  geom_sf_label_repel(
    mapping = aes(label = plot_label)
  ) +
  #coord_sf(crs = st_crs(4326)) +
  #ggtitle("Geodemographic classification of dwellings") +
  xlab("") + ylab("") +
  theme(
    axis.text.x = element_blank(),
    axis.text.y = element_blank(),
    axis.ticks = element_blank(),
    legend.position = "none",
    panel.border = element_rect(colour = "black", fill=NA, size=1)
  ) +
  labs(caption = "Source: Office for National Statistics, Census 2021.\nContains National Statistics data Crown copyright and database right 2022;\nContains Ordnance Survey data Crown copyright and database right 2022.")
```


## Exploring population and households

```{r, warning=FALSE, message=FALSE}
leicester_ggmap <-
  ggmap(
    get_stamenmap(
      c(left = -1.225, bottom = 52.575, right = -1.0, top = 52.7),
      zoom = 14,
      maptype = "toner"
    )
  )

```

```{r}
pop_hh_maps_values <-
  pop_hh_2021 %>% 
  pull(pop_hh_2021) %>% 
  c(
    pop_hh_2011 %>% 
    pull(pop_hh_2011)
  )

pop_hh_maps_breaks <-
  classIntervals(
    c(
      pop_hh_maps_values %>% min() %>% subtract(0.000001),
      pop_hh_maps_values,
      pop_hh_maps_values %>% max() %>% add(0.000001)
    ), 
    n = 9, 
    style = "jenks"
  ) %$%
  brks

pop_hh_maps_cats <-
  pop_hh_maps_values %>% 
  cut(pop_hh_maps_breaks) %>% 
  unique()
```


```{r, echo=FALSE, message=FALSE, warning=FALSE, results='hide', fig.height = 8, fig.width = 8}
leicester_ggmap +
  geom_sf(
    data = 
      pop_hh_2011 %>% 
      mutate(
        pop_hh_2011_breaks =
          cut(
            pop_hh_2011,
            pop_hh_maps_breaks
          )
      ),
    mapping = aes(
      fill = pop_hh_2011_breaks,
      geometry = geometry
    ),
    inherit.aes = FALSE,
    color = NA,
    alpha = 0.8
  ) +
  scale_fill_viridis_d(drop = FALSE) +
  coord_sf(crs = st_crs(4326)) +
  #ggtitle("Geodemographic classification of dwellings") +
  xlab("") + ylab("") +
  theme(
    axis.text.x = element_blank(),
    axis.text.y = element_blank(),
    axis.ticks = element_blank(),
    #legend.position = "none",
    panel.border = element_rect(colour = "black", fill=NA, size=1)
  ) +
  labs(caption = "Source: Office for National Statistics, Census 2011.\nContains National Statistics data Crown copyright and database right 2011;\nContains Ordnance Survey data Crown copyright and database right 2020.\nMap tiles by Stamen Design, under CC BY 3.0. Data by OpenStreetMap, under ODbL.")
```



```{r, echo=FALSE, message=FALSE, warning=FALSE, results='hide', fig.height = 8, fig.width = 8}
leicester_ggmap +
  geom_sf(
    data = 
      pop_hh_2021 %>% 
      mutate(
        pop_hh_2021_breaks =
          cut(
            pop_hh_2021,
            pop_hh_maps_breaks
          )
      ),
    mapping = aes(
      fill = pop_hh_2021_breaks,
      geometry = geometry
    ),
    inherit.aes = FALSE,
    color = NA,
    alpha = 0.8
  ) +
  scale_fill_viridis_d(drop = FALSE) +
  coord_sf(crs = st_crs(4326)) +
  #ggtitle("Geodemographic classification of dwellings") +
  xlab("") + ylab("") +
  theme(
    axis.text.x = element_blank(),
    axis.text.y = element_blank(),
    axis.ticks = element_blank(),
    #legend.position = "none",
    panel.border = element_rect(colour = "black", fill=NA, size=1)
  ) +
  labs(caption = "Source: Office for National Statistics, Census 2021.\nContains National Statistics data Crown copyright and database right 2022;\nContains Ordnance Survey data Crown copyright and database right 2022.\nMap tiles by Stamen Design, under CC BY 3.0. Data by OpenStreetMap, under ODbL.")
```

<!--

### Change 2011 to 2021

```{r}
oa_2011to2021 <-
  read_csv("../storage/Leicester_2011to2021_OAs.csv")
```

-->